---
title: "M8: Forecast Accuracy & Model Performance"
author: "Si Min Loo, Shaochong Xue"
output: pdf_document
always_allow_html: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r package, message=FALSE}
library(readxl)
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(kableExtra)
```

# Loading Data
```{r}
load <- read_excel("tsa-spring-2026-load-forecasting-competition/load.xlsx")
rh <- read_excel("tsa-spring-2026-load-forecasting-competition/relative_humidity.xlsx")
temp <- read_excel("tsa-spring-2026-load-forecasting-competition/temperature.xlsx")

colnames(load)
colnames(temp)
colnames(rh)
```

# Aggregate to Daily
```{r}
# turning into daily data
library(dplyr)

load_daily <- load %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(load = mean(c_across(h1:h24), na.rm = TRUE)) %>%
  ungroup() %>%
  select(date, load)

temp_daily <- temp %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(temp_hourly = mean(c_across(starts_with("t_ws")), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(temp = mean(temp_hourly, na.rm = TRUE))

rh_daily <- rh %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(rh_hourly = mean(c_across(starts_with("rh_ws")), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(rh = mean(rh_hourly, na.rm = TRUE))

data_daily <- load_daily %>%
  left_join(temp_daily, by = "date") %>%
  left_join(rh_daily, by = "date")

data_daily
```

#transform into time series object
```{r}
ts_load_daily <- msts(
  data_daily$load,
  seasonal.periods = c(7, 365.25)
)

```

# Decompose to see the STL 

```{r}
ts_load_daily %>%
  mstl() %>%
  autoplot() +
  labs(title = "STL Decomposition of Daily Electricity Load")

```

# Initial Plots
```{r}
ggplot(data_daily, aes(x = date, y = load)) +
  geom_line() +
  labs(title = "Daily Electricity Load",
       x = "Date",
       y = "Daily Average Load") +
  theme_minimal()
```

# Train and Validation split
```{r}
train <- data_daily %>%
  filter(date >= as.Date("2005-01-01") & date <= as.Date("2010-06-30"))

valid <- data_daily %>%
  filter(date >= as.Date("2010-07-01") & date <= as.Date("2010-07-31"))

nrow(train)
nrow(valid)
head(train)
head(valid)

ts_train <- msts(train$load, seasonal.periods = c(7, 365.25))
ts_valid <- valid$load
h <- length(ts_valid)
```

# Forecasts

## Mean forecast
```{r}
mean_fit <- meanf(ts_train, h = h)
mean_scores <- accuracy(mean_fit$mean, ts_valid)
mean_scores

autoplot(mean_fit) +
  labs(title = "Mean Forecast", y = "Load")
```

## Seasonal forecast
```{r}
snaive_fit <- snaive(ts_train, h = h)
snaive_scores <- accuracy(snaive_fit$mean, ts_valid)
snaive_scores

autoplot(snaive_fit) +
  labs(title = "Seasonal Naive Forecast", y = "Load")
```

## STL + ETS
```{r}
ets_fit <- stlf(ts_train, h = h, method = "ets")
ets_scores <- accuracy(ets_fit$mean, ts_valid)
ets_scores

autoplot(ets_fit) +
  labs(title = "STL + ETS Forecast", y = "Load")

```

## ARIMA + Fourier

```{r}
# # Initial Grid Search for parameters
# K_options <- list(
#   c(1, 6),
#   c(2, 6),
#   c(3, 6),
#   c(1, 12),
#   c(2, 12),
#   c(3, 12),   # max for weekly
#   c(2, 24),
#   c(3, 24)
# )
# 
# results <- lapply(K_options, function(k) {
#   fit <- auto.arima(ts_train, seasonal = FALSE, lambda = 0, #lambda = null
#                     xreg = fourier(ts_train, K = k))
#   fcast <- forecast(fit, xreg = fourier(ts_train, K = k, h = h), h = h)
#   rmse  <- accuracy(fcast$mean, ts_valid)[, "RMSE"]
#   list(K = k, RMSE = rmse, AICc = fit$aicc)
# })
```

  c(2, 24),   # current best

```{r}
arima_fourier_fit <- auto.arima(
  ts_train,
  seasonal = FALSE,
  lambda = 0,
  xreg = fourier(ts_train, K = c(2, 24))
)

arima_fourier_for <- forecast(
  arima_fourier_fit,
  xreg = fourier(ts_train, K = c(2, 24), h = h),
  h = h
)

arima_scores <- accuracy(arima_fourier_for$mean, ts_valid)
arima_scores

autoplot(arima_fourier_for) +
  labs(title = "ARIMA + Fourier Forecast", y = "Load")

```

## SARIMA
```{r}
ts_train_sarima <- ts(train$load, frequency = 7)

sarima_fit <- auto.arima(
  ts_train_sarima,
  seasonal = TRUE,
  lambda = 0
)

sarima_for <- forecast(sarima_fit, h = h)

sarima_scores <- accuracy(sarima_for$mean, ts_valid)
sarima_scores

autoplot(sarima_for) +
  labs(title = "SARIMA Forecast", y = "Load")
```


## TBATS

```{r}
tbats_fit <- tbats(ts_train)
tbats_for <- forecast(tbats_fit, h = h)

tbats_scores <- accuracy(tbats_for$mean, ts_valid)
tbats_scores

autoplot(tbats_for) +
  labs(title = "TBATS Forecast", y = "Load")

```

## NN
```{r}
# p = lagged inputs, P = seasonal lags, size = hidden nodes
nn_fit <- nnetar(ts_train, p = 7, P = 1, size = 10, repeats = 20)
nn_for  <- forecast(nn_fit, h = h)
nn_scores <- accuracy(nn_for$mean, ts_valid)

# Tune: try p=14, P=2 for more memory; increase size for more capacity
nn_fit2 <- nnetar(ts_train, p = 14, P = 2, size = 15, repeats = 20, lambda = 0)
```

## State Space ETS (SSES)
```{r}
sses_fit <- es(ts_train, model = "ZZZ", h = h, holdout = FALSE)
sses_scores <- accuracy(sses_fit$forecast, ts_valid)
sses_scores

plot(sses_fit)
checkresiduals(sses_fit)
```

## State Space BSM (StructTS)
```{r}
# StructTS needs a single-frequency ts object — use weekly (freq=7) like your SARIMA
ts_train_bsm <- ts(train$load, frequency = 7)

# First try letting it find variances freely
ss_fit <- StructTS(ts_train_bsm, type = "BSM", fixed = c(NA, NA, NA, NA))
checkresiduals(ss_fit)

# If convergence issues, fix some variances (uncomment below):
# ss_fit <- StructTS(ts_train_bsm, type = "BSM", fixed = c(0.1, 0.01, 0.3, NA))

ss_for <- forecast(ss_fit, h = h)
ss_scores <- accuracy(ss_for$mean, ts_valid)
ss_scores

autoplot(ss_for) +
  labs(title = "BSM (StructTS) Forecast", y = "Load")
```


# Model Performance, Checking Accuracy
```{r}
# Model Performance, Checking Accuracy
scores <- rbind(
  Mean           = accuracy(mean_fit$mean,         ts_valid)[1,],
  Seasonal_Naive = accuracy(snaive_fit$mean,        ts_valid)[1,],
  STL_ETS        = accuracy(ets_fit$mean,           ts_valid)[1,],
  ARIMA_Fourier  = accuracy(arima_fourier_for$mean, ts_valid)[1,],
  SARIMA         = accuracy(sarima_for$mean,        ts_valid)[1,],
  TBATS          = accuracy(tbats_for$mean,         ts_valid)[1,],
  SSES           = accuracy(sses_fit$forecast,      ts_valid)[1,],
  BSM            = accuracy(ss_for$mean,            ts_valid)[1,]
)

scores_df <- as.data.frame(scores)
scores_df %>%
  kbl(caption = "Forecast Accuracy Comparison") %>%
  kable_styling(full_width = FALSE) %>%
  kable_styling(latex_options = "striped",
                stripe_index = which.min(scores_df$RMSE))  # highlight best

best_model_name <- rownames(scores_df)[which.min(scores_df$RMSE)]
cat("The best model by RMSE is:", best_model_name)
```

# Refit on Best Dataset 

```{r}
# Verify full dataset before refitting
cat("Data range:", as.character(min(data_daily$date)), "to", as.character(max(data_daily$date)), "\n")
cat("Rows:", nrow(data_daily), "\n")

```

```{r}
# Verify full dataset before refitting
cat("Data range:", as.character(min(data_daily$date)), "to", as.character(max(data_daily$date)), "\n")
cat("Rows:", nrow(data_daily), "\n")

ts_full <- msts(data_daily$load, seasonal.periods = c(7, 365.25))
future_h <- 31

# This will take several minutes — normal for TBATS on ~2372 obs
tbats_full_fit <- tbats(ts_full)
tbats_final_forecast <- forecast(tbats_full_fit, h = future_h)

autoplot(tbats_final_forecast) +
  labs(title = "TBATS Final Forecast - July 2011", y = "Daily Load")

# Build submission
template <- read.csv("tsa-spring-2026-load-forecasting-competition/submission_template.csv")
tbats_submission <- template
tbats_submission$load <- as.numeric(tbats_final_forecast$mean)

# Sanity checks
cat("Rows in submission:", nrow(tbats_submission), "\n")        # should be 31
cat("Any NAs:", anyNA(tbats_submission$load), "\n")             # should be FALSE
cat("Load range:", range(tbats_submission$load), "\n")          # should be plausible
print(tbats_submission)

write.csv(tbats_submission, "tbats_submission.csv", row.names = FALSE)
```


# Forecast Results

```{r}
# load template
template <- read.csv("./tsa-spring-2026-load-forecasting-competition/submission_template.csv")
```

## Mean
# ```{r}
# mean_full <- meanf(ts_full, h = future_h)
# 
# mean_for <- template
# mean_for$load <- as.numeric(mean_full$mean)
# 
# mean_for
# ```
# ## SNAIVE
# ```{r}
# snaive_full <- snaive(ts_full, h = future_h)
# 
# snaive_for <- template
# snaive_for$load <- as.numeric(snaive_full$mean)
# 
# snaive_for
# ```


```{r}
write.csv(snaive_for, "initial_result.csv", row.names = FALSE)
```




