---
title: "M8: Forecast Accuracy & Model Performance"
author: "Si Min Loo, Shaochong Xue"
output: pdf_document
always_allow_html: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r package, message=FALSE}
library(readxl)
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(kableExtra)
```

# Loading Data
```{r}
load <- read_excel("tsa-spring-2026-load-forecasting-competition/load.xlsx")
rh <- read_excel("tsa-spring-2026-load-forecasting-competition/relative_humidity.xlsx")
temp <- read_excel("tsa-spring-2026-load-forecasting-competition/temperature.xlsx")

colnames(load)
colnames(temp)
colnames(rh)
```

# Aggregate to Daily
```{r}
# turning into daily data
library(dplyr)

load_daily <- load %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(load = mean(c_across(h1:h24), na.rm = TRUE)) %>%
  ungroup() %>%
  select(date, load)

temp_daily <- temp %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(temp_hourly = mean(c_across(starts_with("t_ws")), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(temp = mean(temp_hourly, na.rm = TRUE))

rh_daily <- rh %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(rh_hourly = mean(c_across(starts_with("rh_ws")), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(rh = mean(rh_hourly, na.rm = TRUE))

data_daily <- load_daily %>%
  left_join(temp_daily, by = "date") %>%
  left_join(rh_daily, by = "date")

data_daily
```

#transform into time series object
```{r}
ts_load_daily <- msts(
  data_daily$load,
  seasonal.periods = c(7, 365.25)
)

```

# Decompose to see the STL 

```{r}
ts_load_daily %>%
  mstl() %>%
  autoplot() +
  labs(title = "STL Decomposition of Daily Electricity Load")

```

# Initial Plots
```{r}
ggplot(data_daily, aes(x = date, y = load)) +
  geom_line() +
  labs(title = "Daily Electricity Load",
       x = "Date",
       y = "Daily Average Load") +
  theme_minimal()
```

# Train and Validation split
```{r}
train <- data_daily %>%
  filter(date >= as.Date("2005-01-01") & date <= as.Date("2010-06-30"))

valid <- data_daily %>%
  filter(date >= as.Date("2010-07-01") & date <= as.Date("2010-07-31"))

nrow(train)
nrow(valid)
head(train)
head(valid)

ts_train <- msts(train$load, seasonal.periods = c(7, 365.25))
ts_valid <- valid$load
h <- length(ts_valid)
```

# Adding exogenous regressors
```{r}
# temp
xreg_train_temp <- as.matrix(train$temp)
xreg_valid_temp <- as.matrix(valid$temp)


dim(xreg_train_temp)
dim(xreg_valid_temp)
head(xreg_train_temp)
head(xreg_valid_temp)
```


# Forecasts

## Mean forecast
```{r}
mean_fit <- meanf(ts_train, h = h)
mean_scores <- accuracy(mean_fit$mean, ts_valid)
mean_scores

autoplot(mean_fit) +
  labs(title = "Mean Forecast", y = "Load")
```

## Seasonal forecast
```{r}
snaive_fit <- snaive(ts_train, h = h)
snaive_scores <- accuracy(snaive_fit$mean, ts_valid)
snaive_scores

autoplot(snaive_fit) +
  labs(title = "Seasonal Naive Forecast", y = "Load")
```

## STL + ETS
```{r}
ets_fit <- stlf(ts_train, h = h, method = "ets")
ets_scores <- accuracy(ets_fit$mean, ts_valid)
ets_scores

autoplot(ets_fit) +
  labs(title = "STL + ETS Forecast", y = "Load")

```

## ARIMA + Fourier

```{r}
# # Initial Grid Search for parameters
# K_options <- list(
#   c(1, 6),
#   c(2, 6),
#   c(3, 6),
#   c(1, 12),
#   c(2, 12),
#   c(3, 12),   # max for weekly
#   c(2, 24),
#   c(3, 24)
# )
# 
# results <- lapply(K_options, function(k) {
#   fit <- auto.arima(ts_train, seasonal = FALSE, lambda = 0, #lambda = null
#                     xreg = fourier(ts_train, K = k))
#   fcast <- forecast(fit, xreg = fourier(ts_train, K = k, h = h), h = h)
#   rmse  <- accuracy(fcast$mean, ts_valid)[, "RMSE"]
#   list(K = k, RMSE = rmse, AICc = fit$aicc)
# })
```

  c(2, 24),   # current best

```{r}
arima_fourier_fit <- auto.arima(
  ts_train,
  seasonal = FALSE,
  lambda = 0,
  xreg = fourier(ts_train, K = c(2, 24))
)

arima_fourier_for <- forecast(
  arima_fourier_fit,
  xreg = fourier(ts_train, K = c(2, 24), h = h),
  h = h
)

arima_scores <- accuracy(arima_fourier_for$mean, ts_valid)
arima_scores

autoplot(arima_fourier_for) +
  labs(title = "ARIMA + Fourier Forecast", y = "Load")

```


# ARIMA + Fourier + Temp
```{r}
# Combine Fourier terms with temperature as xreg
xreg_train_arima_temp <- cbind(
  fourier(ts_train, K = c(2, 24)),
  temp = xreg_train_temp
)

xreg_valid_arima_temp <- cbind(
  fourier(ts_train, K = c(2, 24), h = h),
  temp = xreg_valid_temp
)

arima_temp_fit <- auto.arima(
  ts_train,
  seasonal = FALSE,
  lambda = 0,
  xreg = xreg_train_arima_temp
)

arima_temp_for <- forecast(
  arima_temp_fit,
  xreg = xreg_valid_arima_temp,
  h = h
)

arima_temp_scores <- accuracy(arima_temp_for$mean, ts_valid)
arima_temp_scores

autoplot(arima_temp_for) +
  labs(title = "ARIMA + Fourier + Temperature Forecast", y = "Load")
```

# ARIMA Fourier + Temp + RH
```{r}
xreg_train_arima_weather <- cbind(
  fourier(ts_train, K = c(2, 24)),
  temp = train$temp,
  rh = train$rh
)

xreg_valid_arima_weather <- cbind(
  fourier(ts_train, K = c(2, 24), h = h),
  temp = valid$temp,
  rh = valid$rh
)

arima_weather_fit <- auto.arima(
  ts_train,
  seasonal = FALSE,
  lambda = 0,
  xreg = xreg_train_arima_weather
)

arima_weather_for <- forecast(
  arima_weather_fit,
  xreg = xreg_valid_arima_weather,
  h = h
)

arima_weather_scores <- accuracy(arima_weather_for$mean, ts_valid)
arima_weather_scores
```


## SARIMA
```{r}
ts_train_sarima <- ts(train$load, frequency = 7)

sarima_fit <- auto.arima(
  ts_train_sarima,
  seasonal = TRUE,
  lambda = 0
)

sarima_for <- forecast(sarima_fit, h = h)

sarima_scores <- accuracy(sarima_for$mean, ts_valid)
sarima_scores

autoplot(sarima_for) +
  labs(title = "SARIMA Forecast", y = "Load")
```


## TBATS

```{r}
tbats_fit <- tbats(ts_train)
tbats_for <- forecast(tbats_fit, h = h)

tbats_scores <- accuracy(tbats_for$mean, ts_valid)
tbats_scores

autoplot(tbats_for) +
  labs(title = "TBATS Forecast", y = "Load")

```

## TBATS Cleaned
```{r}
ts_train_clean <- tsclean(ts_train)

tbats_clean_fit <- tbats(ts_train_clean)
tbats_clean_for <- forecast(tbats_clean_fit, h = h)

tbats_clean_scores <- accuracy(tbats_clean_for$mean, ts_valid)
tbats_clean_scores
```
## Recent Window TBATS
```{r}
# Recent-window TBATS experiment
window_results <- data.frame()

start_dates <- as.Date(c(
  "2005-01-01",
  "2006-01-01",
  "2007-01-01",
  "2008-01-01"
))


for (s in start_dates) {
  
  train_window <- train %>%
    filter(date >= s)
  
  ts_train_window <- msts(
    train_window$load,
    seasonal.periods = c(7, 365.25)
  )
  
  fit_window <- tbats(ts_train_window)
  for_window <- forecast(fit_window, h = h)
  
  rmse_window <- accuracy(for_window$mean, ts_valid)[1, "RMSE"]
  
  window_results <- rbind(
    window_results,
    data.frame(
      start_date = as.Date(s), 
      n = nrow(train_window),
      RMSE = rmse_window
    )
  )
}

window_results
```
```{r}
# TBATS using 2007

train_2007 <- train %>%
  filter(date >= as.Date("2007-01-01"))

ts_train_2007 <- msts(
  train_2007$load,
  seasonal.periods = c(7, 365.25)
)

# Fit TBATS
tbats_2007_fit <- tbats(ts_train_2007)

# Forecast validation period
tbats_2007_for <- forecast(
  tbats_2007_fit,
  h = h
)

tbats_2007_scores <- accuracy(
  tbats_2007_for$mean,
  ts_valid
)

tbats_2007_scores

```

# TBATS bias correction
```{r}
tbats_2007_bias_for <- tbats_2007_for$mean + (-35.16369)

accuracy(tbats_2007_bias_for, ts_valid)


```

## TBATS parameter tuning
```{r}
# TBATS detrend
tbats_detrend_fit <- tbats(
  ts_train,
  use.trend = FALSE
)

tbats_detrend_for <- forecast(
  tbats_detrend_fit,
  h = h
)

accuracy(tbats_detrend_for$mean, ts_valid)
```

```{r}
# TBATS with Box-Cox transformation
tbats_boxcox_fit <- tbats(
  ts_train,
  use.box.cox = TRUE
)

tbats_boxcox_for <- forecast(
  tbats_boxcox_fit,
  h = h
)

accuracy(tbats_boxcox_for$mean, ts_valid)
```

```{r}
# TBATS without Box-Cox
tbats_no_boxcox_fit <- tbats(
  ts_train,
  use.box.cox = FALSE
)

tbats_no_boxcox_for <- forecast(
  tbats_no_boxcox_fit,
  h = h
)

accuracy(tbats_no_boxcox_for$mean, ts_valid)
```
```{r}
# TBATS with damped trend
tbats_damped_fit <- tbats(
  ts_train,
  use.trend = TRUE,
  use.damped.trend = TRUE
)

tbats_damped_for <- forecast(
  tbats_damped_fit,
  h = h
)

accuracy(tbats_damped_for$mean, ts_valid)
```



## NN
```{r}
# p = lagged inputs, P = seasonal lags, size = hidden nodes
nn_fit <- nnetar(ts_train, p = 7, P = 1, size = 10, repeats = 20)
nn_for  <- forecast(nn_fit, h = h)
nn_scores <- accuracy(nn_for$mean, ts_valid)

# Tune: try p=14, P=2 for more memory; increase size for more capacity
nn_fit2 <- nnetar(ts_train, p = 14, P = 2, size = 15, repeats = 20, lambda = 0)
```

## State Space ETS (SSES)
```{r}
sses_fit <- es(ts_train, model = "ZZZ", h = h, holdout = FALSE)
sses_scores <- accuracy(sses_fit$forecast, ts_valid)
sses_scores

plot(sses_fit)
checkresiduals(sses_fit)
```

## State Space BSM (StructTS)
```{r}
# StructTS needs a single-frequency ts object — use weekly (freq=7) like your SARIMA
ts_train_bsm <- ts(train$load, frequency = 7)

# First try letting it find variances freely
ss_fit <- StructTS(ts_train_bsm, type = "BSM", fixed = c(NA, NA, NA, NA))
checkresiduals(ss_fit)

# If convergence issues, fix some variances (uncomment below):
# ss_fit <- StructTS(ts_train_bsm, type = "BSM", fixed = c(0.1, 0.01, 0.3, NA))

ss_for <- forecast(ss_fit, h = h)
ss_scores <- accuracy(ss_for$mean, ts_valid)
ss_scores

autoplot(ss_for) +
  labs(title = "BSM (StructTS) Forecast", y = "Load")
```

# Combo (TBAT + ARIMA Fourier)
```{r}
#Refit arima
arima_fourier_full_fit <- auto.arima(
  ts_full,
  seasonal = FALSE,
  lambda = 0,
  xreg = fourier(ts_full, K = c(2, 24))
)

arima_fourier_final_forecast <- forecast(
  arima_fourier_full_fit,
  xreg = fourier(ts_full, K = c(2, 24), h = future_h),
  h = future_h
)


combo_for <- (tbats_for$mean + arima_fourier_for$mean) / 2

combo_scores <- accuracy(combo_for, ts_valid)
combo_scores
```

## Weighted Combo
```{r}
combo_90_10 <- 0.9 * as.numeric(tbats_2007_for$mean) +
               0.1 * as.numeric(tbats_for$mean)

combo_80_20 <- 0.8 * as.numeric(tbats_2007_for$mean) +
               0.2 * as.numeric(tbats_for$mean)

combo_70_30 <- 0.7 * as.numeric(tbats_2007_for$mean) +
               0.3 * as.numeric(tbats_for$mean)

accuracy(combo_90_10, ts_valid)
accuracy(combo_80_20, ts_valid)
accuracy(combo_70_30, ts_valid)
```


# Model Performance, Checking Accuracy
```{r}
# Model Performance, Checking Accuracy
scores <- rbind(
  Mean           = accuracy(mean_fit$mean,         ts_valid)[1,],
  Seasonal_Naive = accuracy(snaive_fit$mean,        ts_valid)[1,],
  STL_ETS        = accuracy(ets_fit$mean,           ts_valid)[1,],
  ARIMA_Fourier  = accuracy(arima_fourier_for$mean, ts_valid)[1,],
  ARIMA_Fourier_Temp = accuracy(arima_temp_for$mean,ts_valid)[1,],
  ARIMA_Fourier_Weather = accuracy(arima_weather_for$mean, ts_valid)[1,],
  SARIMA         = accuracy(sarima_for$mean,        ts_valid)[1,],
  TBATS          = accuracy(tbats_for$mean,         ts_valid)[1,],
  TBATS_Clean    = accuracy(tbats_clean_for$mean,   ts_valid)[1,],
  TBATS_2007     = accuracy(tbats_2007_for$mean,    ts_valid)[1,],
  TBATS_Detrend       = accuracy(tbats_detrend_for$mean, ts_valid)[1,],
  TBATS_BoxCox         = accuracy(tbats_boxcox_for$mean, ts_valid)[1,],
  TBATS_No_BoxCox      = accuracy(tbats_no_boxcox_for$mean, ts_valid)[1,],
  TBATS_Damped         = accuracy(tbats_damped_for$mean, ts_valid)[1,],
  SSES           = accuracy(sses_fit$forecast,      ts_valid)[1,],
  BSM            = accuracy(ss_for$mean,            ts_valid)[1,],
  Combo          = accuracy(combo_for,              ts_valid)[1,]
)

scores_df <- as.data.frame(scores)
scores_df %>%
  kbl(caption = "Forecast Accuracy Comparison") %>%
  kable_styling(full_width = FALSE) %>%
  kable_styling(latex_options = "striped",
                stripe_index = which.min(scores_df$RMSE))  # highlight best

# top 5 models
scores_ranked <- scores_df[order(scores_df$RMSE), ]
top5 <- head(scores_ranked, 5)

cat("Top 5 models by RMSE:\n")
for (i in 1:5) {
  cat(sprintf("  #%d: %-25s RMSE = %.4f\n", i, rownames(top5)[i], top5$RMSE[i]))
}
```

# Refit on Best Dataset 

```{r}
# Verify full dataset before refitting
cat("Data range:", as.character(min(data_daily$date)), "to", as.character(max(data_daily$date)), "\n")
cat("Rows:", nrow(data_daily), "\n")

```

```{r}
# Verify full dataset before refitting
cat("Data range:", as.character(min(data_daily$date)), "to", as.character(max(data_daily$date)), "\n")
cat("Rows:", nrow(data_daily), "\n")

ts_full <- msts(data_daily$load, seasonal.periods = c(7, 365.25))
future_h <- 31

# This will take several minutes — normal for TBATS on ~2372 obs
tbats_full_fit <- tbats(ts_full)
tbats_final_forecast <- forecast(tbats_full_fit, h = future_h)

autoplot(tbats_final_forecast) +
  labs(title = "TBATS Final Forecast - July 2011", y = "Daily Load")

# Build submission
template <- read.csv("tsa-spring-2026-load-forecasting-competition/submission_template.csv")
tbats_submission <- template
tbats_submission$load <- as.numeric(tbats_final_forecast$mean)

# Sanity checks
cat("Rows in submission:", nrow(tbats_submission), "\n")        # should be 31
cat("Any NAs:", anyNA(tbats_submission$load), "\n")             # should be FALSE
cat("Load range:", range(tbats_submission$load), "\n")          # should be plausible
print(tbats_submission)

write.csv(tbats_submission, "tbats_submission.csv", row.names = FALSE)
```


# Forecast Results

```{r}
# load template
template <- read.csv("./tsa-spring-2026-load-forecasting-competition/submission_template.csv")
```

## Mean
# ```{r}
# mean_full <- meanf(ts_full, h = future_h)
# 
# mean_for <- template
# mean_for$load <- as.numeric(mean_full$mean)
# 
# mean_for
# ```
# ## SNAIVE
# ```{r}
# snaive_full <- snaive(ts_full, h = future_h)
# 
# snaive_for <- template
# snaive_for$load <- as.numeric(snaive_full$mean)
# 
# snaive_for
# ```


```{r}
write.csv(snaive_for, "initial_result.csv", row.names = FALSE)
```



```{r}
# output combine result
# Build submission
combo_submission <- template
combo_submission$load <- as.numeric(combo_for)

# Sanity checks
cat("Rows in submission:", nrow(combo_submission), "\n")
cat("Any NAs:", anyNA(combo_submission$load), "\n")
cat("Load range:", range(combo_submission$load), "\n")
print(combo_submission)

write.csv(combo_submission, "combo_submission.csv", row.names = FALSE)
```

## Output cleaned tbats
```{r}
# Clean full data
ts_full_clean <- tsclean(ts_full)

# Fit TBATS on cleaned full data
tbats_clean_full_fit <- tbats(ts_full_clean)

# Forecast
tbats_clean_final_forecast <- forecast(tbats_clean_full_fit, h = 31)
```

```{r}
# output TBATS + tsclean result

# Build submission
tbats_clean_submission <- template
tbats_clean_submission$load <- as.numeric(tbats_clean_final_forecast$mean)

# Sanity checks
cat("Rows in submission:", nrow(tbats_clean_submission), "\n")
cat("Any NAs:", anyNA(tbats_clean_submission$load), "\n")
cat("Load range:", range(tbats_clean_submission$load), "\n")
print(tbats_clean_submission)

write.csv(tbats_clean_submission, "tbats_clean_submission.csv", row.names = FALSE)
```

## Output 2007+ TBATS
```{r}
data_2007 <- data_daily %>%
  filter(date >= as.Date("2007-01-01"))

ts_2007 <- msts(
  data_2007$load,
  seasonal.periods = c(7, 365.25)
)

tbats_2007_final_fit <- tbats(ts_2007)

tbats_2007_forecast <- forecast(
  tbats_2007_final_fit,
  h = 31
)
```

```{r}

# Build submission
tbats_2007_submission <- template
tbats_2007_submission$load <- as.numeric(tbats_2007_forecast$mean)

# Sanity checks
cat("Rows in submission:", nrow(tbats_2007_submission), "\n")
cat("Any NAs:", anyNA(tbats_2007_submission$load), "\n")
cat("Load range:", range(tbats_2007_submission$load), "\n")
print(tbats_2007_submission)

write.csv(tbats_2007_submission, "tbats_2007_submission.csv", row.names = FALSE)
```

# Output TBATS bias correction
```{r}
tbats_2007_bias_forecast <- tbats_2007_forecast$mean + (-35.16369)
```

```{r}
# output TBATS recent 2007 with bias correction result

# Build submission
tbats_2007_bias_submission <- template
tbats_2007_bias_submission$load <- as.numeric(tbats_2007_bias_forecast)

# Sanity checks
cat("Rows in submission:", nrow(tbats_2007_bias_submission), "\n")
cat("Any NAs:", anyNA(tbats_2007_bias_submission$load), "\n")
cat("Load range:", range(tbats_2007_bias_submission$load), "\n")
print(tbats_2007_bias_submission)

write.csv(tbats_2007_bias_submission, 
          "tbats_2007_bias_submission.csv", 
          row.names = FALSE)
```

# Output TBATS NO BOXCOX
```{r}
# Final TBATS without Box-Cox on full data

tbats_no_boxcox_final_fit <- tbats(
  ts_full,
  use.box.cox = FALSE
)

tbats_no_boxcox_final_forecast <- forecast(
  tbats_no_boxcox_final_fit,
  h = 31
)
```

```{r}
# output TBATS_No_BoxCox result

# Build submission
tbats_no_boxcox_submission <- template
tbats_no_boxcox_submission$load <- as.numeric(tbats_no_boxcox_final_forecast$mean)

# Sanity checks
cat("Rows in submission:", nrow(tbats_no_boxcox_submission), "\n")
cat("Any NAs:", anyNA(tbats_no_boxcox_submission$load), "\n")
cat("Load range:", range(tbats_no_boxcox_submission$load), "\n")
print(tbats_no_boxcox_submission)

write.csv(tbats_no_boxcox_submission, 
          "tbats_no_boxcox_submission.csv", 
          row.names = FALSE)
```



