---
title: "M8: Forecast Accuracy & Model Performance"
author: "Si Min Loo, Shaochong Xue"
output: pdf_document
always_allow_html: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 80), tidy = FALSE)
```

GitHub Repository: https://github.com/AllenX0627/LooXue_ENV797_TSA_ForecastCompetition_S26.git

AI clarification: In this project, AI tools were used to assist with debugging and document formatting.

```{r packages, message=FALSE, warning=FALSE}
library(readxl)
library(lubridate)
library(dplyr)
library(ggplot2)
library(forecast)
library(tseries)
library(tidyverse)
library(kableExtra)
```

# Loading Data

```{r load-data}
load <- read_excel("tsa-spring-2026-load-forecasting-competition/load.xlsx")
rh   <- read_excel("tsa-spring-2026-load-forecasting-competition/relative_humidity.xlsx")
temp <- read_excel("tsa-spring-2026-load-forecasting-competition/temperature.xlsx")
```

# Data Wrangling: Aggregate to Daily

```{r wrangle}
load_daily <- load %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(load = mean(c_across(h1:h24), na.rm = TRUE)) %>%
  ungroup() %>%
  select(date, load)

temp_daily <- temp %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(temp_hourly = mean(c_across(starts_with("t_ws")), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(temp = mean(temp_hourly, na.rm = TRUE))

rh_daily <- rh %>%
  mutate(date = as.Date(date)) %>%
  rowwise() %>%
  mutate(rh_hourly = mean(c_across(starts_with("rh_ws")), na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(rh = mean(rh_hourly, na.rm = TRUE))

data_daily <- load_daily %>%
  left_join(temp_daily, by = "date") %>%
  left_join(rh_daily,  by = "date")

head(data_daily)
```

# Exploratory Analysis

```{r eda-plot}
ggplot(data_daily, aes(x = date, y = load)) +
  geom_line() +
  labs(title = "Daily Electricity Load",
       x = "Date", y = "Daily Average Load") +
  theme_minimal()
```

```{r stl-decomp}
ts_load_daily <- msts(data_daily$load, seasonal.periods = c(7, 365.25))

ts_load_daily %>%
  mstl() %>%
  autoplot() +
  labs(title = "STL Decomposition of Daily Electricity Load")
```

# Train / Validation Split

```{r train-valid-split}
train <- data_daily %>%
  filter(date >= as.Date("2005-01-01") & date <= as.Date("2010-06-30"))

valid <- data_daily %>%
  filter(date >= as.Date("2010-07-01") & date <= as.Date("2010-07-31"))

cat("Training rows:", nrow(train), "\n")
cat("Validation rows:", nrow(valid), "\n")

ts_train <- msts(train$load, seasonal.periods = c(7, 365.25))
ts_valid <- valid$load
h        <- length(ts_valid)
```

---

# Top 5 Models

## Model 1: TBATS

TBATS fitted on the full training window (2005–2010).

```{r tbats}
tbats_fit <- tbats(ts_train)
tbats_for <- forecast(tbats_fit, h = h)

tbats_scores <- accuracy(tbats_for$mean, ts_valid)
tbats_scores

autoplot(tbats_for) +
  labs(title = "TBATS Forecast", y = "Load")
```

## Model 2: TBATS_Clean

TBATS fitted after removing outliers with `tsclean()`.

```{r tbats-clean}
ts_train_clean  <- tsclean(ts_train)

tbats_clean_fit <- tbats(ts_train_clean)
tbats_clean_for <- forecast(tbats_clean_fit, h = h)

tbats_clean_scores <- accuracy(tbats_clean_for$mean, ts_valid)
tbats_clean_scores

autoplot(tbats_clean_for) +
  labs(title = "TBATS Clean Forecast", y = "Load")
```

## Model 1: TBATS (No Box-Cox)

TBATS fitted on the full training window with Box-Cox transformation disabled.

```{r tbats-no-boxcox}
tbats_no_boxcox_fit <- tbats(ts_train, use.box.cox = FALSE)
tbats_no_boxcox_for <- forecast(tbats_no_boxcox_fit, h = h)

tbats_no_boxcox_scores <- accuracy(tbats_no_boxcox_for$mean, ts_valid)
tbats_no_boxcox_scores

autoplot(tbats_no_boxcox_for) +
  labs(title = "TBATS (No Box-Cox) Forecast", y = "Load")
```

## Model 4: Combo (TBATS + ARIMA Fourier)

Simple average ensemble of TBATS and ARIMA+Fourier forecasts.

```{r arima-fourier-for-combo}
arima_fourier_fit <- auto.arima(
  ts_train,
  seasonal = FALSE,
  lambda   = 0,
  xreg     = fourier(ts_train, K = c(2, 24))
)

arima_fourier_for <- forecast(
  arima_fourier_fit,
  xreg = fourier(ts_train, K = c(2, 24), h = h),
  h    = h
)
```

```{r combo}
combo_for    <- (tbats_for$mean + arima_fourier_for$mean) / 2
combo_scores <- accuracy(combo_for, ts_valid)
combo_scores
```

## Model 5: ARIMA + Fourier

ARIMA with Fourier terms (K = c(2, 24)) for dual seasonality.

```{r arima-fourier}
arima_scores <- accuracy(arima_fourier_for$mean, ts_valid)
arima_scores

autoplot(arima_fourier_for) +
  labs(title = "ARIMA + Fourier Forecast (K = c(2,24))", y = "Load")
```

---

# Model Performance Summary

```{r accuracy-summary}
scores <- rbind(
  TBATS         = tbats_scores[1, ],
  TBATS_Clean   = tbats_clean_scores[1, ],
  TBATS_No_BoxCox = tbats_no_boxcox_scores[1, ],
  Combo         = combo_scores[1, ],
  ARIMA_Fourier = arima_scores[1, ]
)

scores_df      <- as.data.frame(scores)
scores_ranked  <- scores_df[order(scores_df$RMSE), ]

scores_ranked %>%
  kbl(caption = "Top 5 Model Accuracy (Validation: July 2010)",
      digits = 4) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("striped", "hold_position"),
                stripe_index  = 1)

cat("\nTop 5 models by RMSE:\n")
for (i in seq_len(nrow(scores_ranked))) {
  cat(sprintf("  #%d: %-25s RMSE = %.4f\n",
              i, rownames(scores_ranked)[i], scores_ranked$RMSE[i]))
}
```
